The Multi-Stencil Language

### 2016-01-25 ARTICLES/IJPP/article/formalism.tex
article IJPP corrections formalism et parallelism Helene 25/01/2016

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item An entity $e$ of a mesh $\mathcal{M}=(V,E)$ is defined as a subset of its vertices and edges, $e\subset (V\cup E)$.
\item A mesh entities group is denoted $E$ such that $E \in \mathcal{P}(V\cup E)$. It represents a set of entities of the same type.
\item Finally the set of mesh entities groups of a simulation is denoted $\mathcal{E}$.
\end{itemize}
%\end{mydef}

@@

In a numerical simulation, as a set of scalars can be applied onto a mesh with a nil dimension, a set of data elements, or quantities, can be applied onto meshes of dimension superior or equal to zero. Those quantities represent, as well as scalars, the set of values to compute, or to use, for computations.

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item \textit{A quantity (or data) is a function $\delta: E_{\delta} \mapsto V_{\delta}$ which associates each entity of a group $e\in E_{\delta}$ to a value $v\in V_{\delta}$ where $V_{\delta}$ is a data type, typically $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.
\item The set of quantities applied onto the mesh is denoted $\Delta$.
\item In the rest of this paper, the group of mesh entities on which a quantity $\delta$ is mapped is denoted $entity(\delta)=E_{\delta}$.}
\end{itemize}
%\end{mydef}

@@

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A computation domain $D$ is a subpart of a mesh entities group, $D \subset E \in \mathcal{E}$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item A neighborhood $n$ is a function which for a given entity $e \in E_i$, returns a set of $m$ entities in $E_j$, $n : E_i \rightarrow E_j^m$. One can notice that $i = j$ is possible. Most of the time, such a neighborhood is called a stencil shape.
\item The set of neihborhood functions in a numerical simulation is denoted $\mathcal{N}$.
\end{itemize}

@@

\noindent \textbf{Property}
Considering a reduction kernel $k(S,R,(w,D),comp)$, $comp$ must be a binary and associative operation on the type $V$, $comp: V \times V \rightarrow V$.

\noindent \textbf{Definitions}
\begin{itemize}
\item The set of $n$ ordered computation kernels of a numerical simulation is denoted $\Gamma = [k_i]_{0 \leq i \leq n-1}$, such that $\forall k_i,k_j \in \Gamma$, if $i \leq j$, then $k_i$ is computed before $k_j$.
\item Finally, a \textit{multi-stencil program} is defined by the septuplet 
\begin{equation*}
\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta, \mathcal{S},\Gamma)
\end{equation*}
\end{itemize}

>>>>

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item An entity $e$ of a mesh $\mathcal{M}=(V,E)$ is defined as a subset of its vertices and edges, $e\subset V\cup E$.
\item A mesh entities group is denoted $E$ such that $E \in \mathcal{P}(V\cup E)$. It represents a set of entities of the same type.
\item The set of mesh entities groups of a simulation is denoted $\mathcal{E}$.
\item Finally, the mesh on which is applied a group of mesh entities $E$ is denoted $mesh(E)$
\end{itemize}
%\end{mydef}

@@

In a numerical simulation, as a set of scalars can be applied onto a mesh with a nil dimension, a set of data elements, or quantities, can be applied onto meshes of dimension superior to zero. Those quantities represent, as well as scalars, the set of values to compute, or to use, for computations.

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A quantity (or data) is a function $\delta: E_{\delta} \mapsto V_{\delta}$ which associates each entity of a group $e\in E_{\delta}$ to a value $v\in V_{\delta}$ where $V_{\delta}$ is a data type, typically $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.
\item The set of quantities applied onto the mesh is denoted $\Delta$.
\item In the rest of this paper, the group of mesh entities on which a quantity $\delta$ is mapped is denoted $entity(\delta)=E_{\delta}$.
\end{itemize}
%\end{mydef}

@@

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A computation domain $D$ is a subpart of a mesh entities group, $D \subset E \in \mathcal{E}$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item A neighborhood $n$ is a function which for a given entity $e \in E_i$, returns a set of $m$ entities in $E_j$, $n : E_i \rightarrow E_j^m$. One can notice that $i = j$ is possible. Most of the time, such a neighborhood is called a stencil shape.
\item The set of neihborhood functions in a numerical simulation is denoted $\mathcal{N}$.
\end{itemize}

@@

\medskip
\noindent \textbf{Property}
Considering a reduction kernel $k(S,R,(w,D),comp)$, $comp$ must be a binary and associative operation on the type $V$, $comp: V \times V \rightarrow V$.

\medskip
\noindent \textbf{Definitions}
\begin{itemize}
\item The set of $n$ ordered computation kernels of a numerical simulation is denoted $\Gamma = [k_i]_{0 \leq i \leq n-1}$, such that $\forall k_i,k_j \in \Gamma$, if $i \leq j$, then $k_i$ is computed before $k_j$.
\item Finally, a \textit{multi-stencil program} is defined by the septuplet 
\begin{equation*}
\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta, \mathcal{S},\Gamma)
\end{equation*}
\end{itemize}

### 2016-01-25 ARTICLES/IJPP/article/parallelism.tex
article IJPP corrections formalism et parallelism Helene 25/01/2016

\begin{mydef}
A synchronization between two computations $k_i \pprec k_j$ is defined as a specific computation 
\begin{equation*}
k_{sync}(S,R,w,comp)
\end{equation*}
\end{mydef}
where $S=\emptyset$, $R=\{(w_i,n_j)\}$, $w=(w_i,n_j)$ and $comp=identity$. In other words, $w_i$ has to be synchronized for the neighborhood $n_j$.

@@

When a data parallelism is applied, the other type of computation which is responsible for additional synchronizations is the reduction. Actually, the reduction is applied locally on each subset of entities, on each resource, which results in $p$ (number of resources) scalar value, one on each resource. Thus, to perform the final reduction a synchronization is needed to get the $p$ scalars onto a single resource, to finally get the reduction scalar.

\begin{mydef}
For a reduction kernel $k(S,R,(w,D),comp)$, if we denote by $w^i$, $0 \leq i<p$, the local $w$ computed on the resource $i$, a reduction synchronization is defined as the specific computation 
\begin{equation*}
k_{sync}(S,R_s,w,comp)
\end{equation*}
\end{mydef}
where, $S=\emptyset$, $R=\{(w^0,entity(w^0)) \dots (w^{p-1},entity(w^{p-1}))\}$ and $w$ and $comp$ are the same than the actual reduction.

@@

\begin{mydef}
The concatenation of two ordered lists of respectively $n$ and $m$ computations $l_1=[k_i]_{0 \leq i \leq n-1}$ and $l_2=[c'_i]_{0 \leq i \leq m-1}$ is denoted $l_1 \cdot l_2$ and is equal to a new ordered list $l_3=[c_0,\dots,c_{n-1},c'_0,\dots,c'_{m-1}]$.
\end{mydef}

\begin{mydef}
From the ordered list of computation $\Gamma$, a new ordered list $\Gamma_{sync}$ is obtained from the call $\Gamma_{sync} = F_{sync}(\Gamma,0)$, where $F_{sync}$ is the recursive function defined as
\begin{equation*}
F_{sync}(\Gamma,j) = 
\begin{cases} 	[] \mbox{ if }j=|\Gamma|\\
				[k_{sync}(\emptyset,\{w_i\},w_i,identity,\emptyset), k_j] \cdot F_{sync}(\Gamma,j+1)\\
				\text{ } \qquad \forall i<j \text{, and }k_i \pprec k_j\\
				[k_j, k_{sync}(\emptyset,,w_j,comp_j)] \cdot F_{sync}(\Gamma,j+1)\\ 
				\text{ } \qquad \mbox{if } w_j \in \mathcal{S} \text{, and } R_j=\{(r_j,entity(r_j))\}\\
				[\Gamma[j]] \cdot F_{sync}(\Gamma,j+1) \text{ else}.
\end{cases}
\end{equation*}
\end{mydef}

@@

\begin{mydef}
From the ordered list $\Gamma_{sync}$ of computation kernels, a directed dependency graph $\Gamma_{dep}(V,A)$ is obtained from the call $T_{dep}(\Gamma_{sync},0)$, where $F_{dep}$ is the recursive function
\begin{equation*}
F_{dep}(\Gamma_{sync},j) = 
\begin{cases} 	(\{\},\{\}) \mbox{ if }j=|\Gamma_{sync}|\\
				(k_j, \{(\overset{\frown}{k_i,k_j})\mbox{, }\forall i < j \mbox{, } k_i\prec k_j \})\\
				\text{ } \qquad \cup F_{dep}(\Gamma_{sync},j+1) \mbox{ if }j<|\Gamma_{sync}|
\end{cases}
\end{equation*}
\end{mydef}

@@

As a result of the hybrid parallelization, each resource $0 \leq m \leq p-1$ perform a multi-stencil program, defined by
\begin{equation*}
\mathcal{MSP}_m(T,\mathcal{M}_m,\mathcal{E}_m,\mathcal{D}_m,\Delta_m,\Gamma_{dep}).
\end{equation*}
The set of computations $\Gamma_{dep}$ is a dependency graph between computations $k_i$ of $\Gamma$ and the synchronization kernels added into $\Gamma_{sync}$. $\Gamma_{dep}$ can be built from the call to 
\begin{equation*}
F_{dep}(F_{sync}(\Gamma,0),0).
\end{equation*}

>>>>

\begin{mydef}
A synchronization between two computations $k_i \pprec k_j$ is defined as a specific computation 
\begin{equation*}
k_{sync}(S,R,(w,D),comp)
\end{equation*}
\end{mydef}
where $S=\emptyset$, $R=\{(w_i,\bigcup_{e \in entity(w_i)} n_j(e))\}$, $w=(w_i,\bigcup_{e \in entity(w_i)} n_j(e)))$ and $comp=identity$. In other words, $w_i$ has to be synchronized for the neighborhood $n_j$ for all entities of $w_i$.

@@

When data parallelism is applied, the other type of computation which is responsible for additional synchronizations is the reduction. Actually, the reduction is applied locally on each subset of entities, on each resource, which results in $p$ (number of resources) scalar values, one on each resource. Thus, to perform the final reduction a synchronization is needed to get the $p$ scalars onto a single resource, to finally get the reduction scalar.

\begin{mydef}
For a reduction kernel $k(S,R,(w,D),comp)$, if we denote by $w^i$, $0 \leq i<p$, the local $w$ computed on the resource $i$, a reduction synchronization is defined as the specific computation 
\begin{equation*}
k_{sync}(S,R_s,(w,D),comp)
\end{equation*}
\end{mydef}
where, $S=\emptyset$, $R_s=\{(w^0,entity(w^0)) \dots (w^{p-1},entity(w^{p-1}))\}$, and $w$, $D=entity(w)$ and $comp$ are the same than the actual reduction kernel $k$.

@@

\begin{mydef}
The concatenation of two ordered lists of respectively $n$ and $m$ computations $l_1=[k_i]_{0 \leq i \leq n-1}$ and $l_2=[k'_i]_{0 \leq i \leq m-1}$ is denoted $l_1 \cdot l_2$ and is equal to a new ordered list $l_3=[k_0,\dots,k_{n-1},k'_0,\dots,k'_{m-1}]$.
\end{mydef}

\begin{mydef}
From the ordered list of computation $\Gamma$, a new ordered list $\Gamma_{sync}$ is obtained from the call $\Gamma_{sync} = F_{sync}(\Gamma,0)$, where $F_{sync}$ is the recursive function defined as
\begin{equation*}
F_{sync}(\Gamma,j) = 
\begin{cases} 	\bullet [] \mbox{ if }j=|\Gamma|\\
				\bullet [k_{sync}(\emptyset,\\
				\{(w_i,\bigcup_{e \in entity(w_i)} n_j(e))\},\\
				(w_i,\bigcup_{e \in entity(w_i)} n_j(e))),\\
				identity,\emptyset), k_j] \cdot F_{sync}(\Gamma,j+1)\\
				\text{ } \qquad \forall i<j \text{, and }k_i \pprec k_j\\
				\bullet [k_j, k_{sync}(\emptyset,\\
				\{(w^0_j,entity(w^0_j)) \dots (w^{p-1}_j,entity(w^{p-1}_j))\},\\
				(w_j,entity(w_j)),comp_j)] \cdot F_{sync}(\Gamma,j+1)\\ 
				\text{ } \qquad \mbox{if } w_j \in \mathcal{S} \text{, and } R_j=\{(r_j,entity(r_j))\}\\
				\bullet [\Gamma[j]] \cdot F_{sync}(\Gamma,j+1) \text{ else}.
\end{cases}
\end{equation*}
\end{mydef}

@@

\begin{mydef}
From the ordered list $\Gamma_{sync}$ of computation kernels, a directed dependency graph $\Gamma_{dep}(V,A)$ is obtained from the call $F_{dep}(\Gamma_{sync},0)$, where $F_{dep}$ is the recursive function
\begin{equation*}
F_{dep}(\Gamma_{sync},j) = 
\begin{cases} 	\bullet (\{\},\{\}) \mbox{ if }j=|\Gamma_{sync}|\\
				\bullet (k_j, \{(\overset{\frown}{k_i,k_j})\mbox{, }\forall i < j \mbox{, } k_i\prec k_j \})\\
				\text{ } \qquad \cup F_{dep}(\Gamma_{sync},j+1) \mbox{ if }j<|\Gamma_{sync}|
\end{cases}
\end{equation*}
\end{mydef}

@@

As a result of the hybrid parallelization, each resource $0 \leq m \leq p-1$ perform a multi-stencil program, defined by
\begin{equation*}
\mathcal{MSP}_m(T,\mathcal{M}_m,\mathcal{E}_m,\mathcal{D}_m,\Delta_m,\Gamma_{dep}).
\end{equation*}
The set of computations $\Gamma_{dep}$ is a dependency graph between computations $k_i$ of $\Gamma$ and synchronizations of kernels added into $\Gamma_{sync}$. $\Gamma_{dep}$ can be built from the call to 
\begin{equation*}
F_{dep}(F_{sync}(\Gamma,0),0).
\end{equation*}
